Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-30 MScDiplom_Lemenkova.tex
added Section 'Radiative Transfer Models (RTM)' - comment 5 of Valentijn

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\section*{}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\setcounter{lofdepth}{2}
\pagebreak

@@

\subsection{\textit{In-situ} observations of the seagrass meadows}
The traditional methods of in-situ seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03} \label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, markers, GPS and other equipment. The seagrass sampling is
taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}. \\
Other methods of seagrass \textit{in-situ} monitoring are based on
using the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then
collect the return echoes for the analysis, yet this requires specialized equipment and is mostly used in
the depth waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that they are not effective at identifying the composition or even the presence of
submerged aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}, because
initially acoustic methods were intended for the bathymetric surveying, i.e. they are not adjusted for
the benthic habitat discrimination. The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated GIS approach, as, for example, reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.

@@

\subsection[Application of remote sensing data]{Application of remote sensing data towards seagrass mapping}
Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, GPS data, remote sensing and SCUBA-based fieldwork measurements.
SCUBA-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-11.jpg}
	\caption{In situ optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:13}
\end{wrapfigure}

The underwater videography with a GPS is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}. However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. Integrated together with GIS vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.\\
However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Mellors09a}\label{Mellors09a}.
Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. \\
Spectral measurements of the target objects are made
by means of the radiometers, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

@@

Studies of spectral reflectance of the different seagrass species comparing to the spectra of sand and
other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (LAI) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.
Spectral reflectance of the seagrass  (Fig.\ref{fig:13}) is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason for marine mapping of submerged vegetation
and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).

\begin{table}[htbp]
	\caption{\textbf{Characteristics of selected ocean-colour sensors}}
	\begin{center}
		\rowcolors{1}{OldLace}{FloralWhite}
		\begin{tabular}{|p{2cm}|p{20mm}|p{15mm}|p{17mm}|p{15mm}|p{15mm}|p{15mm}|}
			\hline\hline
			\textbf{Satellite Sensor:} & {\textbf{Landsat}} & {\textbf{SPOT}} & {\textbf{ENVISAT}} & {\textbf{Terra}} & {\textbf{GeoEye}} & {\textbf{Nimbus 7}} \\ \hline\hline
				\textbf{Origin} & USA & France & Europe (ESA) & USA (NASA) & USA & USA \\ \hline
				\textbf{Instrument} & \textit{TM, ETM+} & \textit{Pan, XS} & \textit{MERIS} & \textit{MODIS} & \textit{SeaWiFS} & \textit{CZCS}\\ \hline
			\textbf{Number of channels} & 8 & 6 & 15 & 36 & 8 & 5 \\ \hline
			\textbf{Wavelength coverage} & 185-1700 & 500-1750 & 412.5- 900 & 469-1640 & 412-865 & 443-750 \\ \hline
			\textbf{Ground resolution (nadir), m} & 15-pan,30(MS),60 (TIR) & 2.5, 5, 10,20 & 1.2 km/300 m & 1.0 km & 1.13 km & 825 m \\ \hline
			\textbf{Band Width (nm)} & 0.45-10.4 & 10,20 & 2.5,7.5,9,10 14,20 & 0.18,0.3,0.5,10 - 50 & 20,40 & 20,100 \\ \hline
			\textbf{Launched} & 1972, 1999 & 1986 & 2002 & 1999 & 1997 & 1978 \\ \hline
			\textbf{Recurrent period, days} & 16 & 24 & 35 & 16 & 16 & 16 \\ \hline
		\end{tabular}
	\end{center}
	\label{tab:1}
\end{table}

In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.

@@

The Landsat TM and Landsat ETM+ data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
Landsat TM towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.
Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}.

@@

\subsection{Fieldwork data collection}
Seagrass sampling has been performed at two research stations at Crete Island - Ligaria beach
(36°20′N 22°59′E) and Xerocampos (35°12′N 26°18′E), at depths lesser than 4 meters.

>>>>

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\section*{}
\setcounter{lofdepth}{2}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\pagebreak

@@

@@

\subsection[Radiative Transfer Models (RTM)]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing tools}

Understanding radiance distribution within a water column is necessary for the studies of the underwater visibility, because scattering properties of the water body naturally vary with changing depth, wavelength and environmental conditions.  Simulation of the radiance quantities for natural water bodies enables to analyse seafloor color remote sensing properties. Therefore, the artificial modelling of the seawater optical properties by means of the Radiative Transfer Modelling (further RTM) is used when water optics is studied under changing environment (e.g., depths, sun angle, suspended particles in water column). \\ In such cases a retrieval of water optical parameters from the remote measurements should be tested and analyzed. \\
We briefly discuss below the most effective RTM software and algorithms from a range of various best-known up-to-date tools. 
	\begin{itemize}
		\item [*] The \textit{Second Simulation of a Satellite Signal in the Solar Spectrum}, version 1 \href{http://6s.ltdri.org/}{(6SV1)} is a US atmospheric correction algorithm, developed in the University of Maryland; adjusted for the \href{http://modis.gsfc.nasa.gov/}{NASA MODIS} satellite imagery lookup tables. The 6S algorithm has been implemented by the \href{http://grass.osgeo.org/grass64/manuals/html64_user/i.atcorr.html}{GRASS} software for the atmospheric correction 										 
		\item [*] Another example of the RTM adjusted for a specific imagery is a German \href{http://www.imk-asf.kit.edu/english/312.php}{\textit{KOPRA RTM}}, fitted for \href{http://envisat.esa.int/instruments/meris/}{MIPAS/ENVISAT} imagery.
		\item [*]  The \href{http://www.sequoiasci.com/products/Hydrolight.aspx}{\textit{HydroLight RTM}} commercial software is an advanced model for oceanic radiative transfer calculations, developed by the Sequoia Scientific, Inc., USA. It is designed to solve a wide range of problems in optical oceanography and limnology.   
		\item [*] The British \href{http://research.metoffice.gov.uk/research/interproj/nwpsaf/rtm/}{\textit{RTTOV-9 RTM}} is a radiative transfer model, developed for nadir viewing atmospheric sounders and imagers. It includes a number of useful tolls, e.g. can compute sea-surface emissivity for each channel, enables, to specify cloudiness for radiance calculations, etc.  This program runs under Unix/Linux platforms and is open source. 
		\item [*] The \href{http://rtweb.aer.com/}{\textit{Rapid Radiative Transfer Model (RRTM)}} is one more RTM freeware, designed in the USA, Massachusetts. 
		\item [*] The \href{http://www.modtran.org/}{\textit{MODTRAN}}, a commercial software from Spectral Sciences Inc., is another example of the RTM. Among its newest updates is generation of atmospheric correction data; it also has an option to write spectral output in binary, and convert it to ASCII.
		\item [*] A \textit{Discrete Ordinates Radiative Transfer Program for a Multi-Layered Plane-Parallel Medium} \href{ftp://climate1.gsfc.nasa.gov/wiscombe/Multiple_Scatt/DISORT1.2/}{(DISORT)}  is another RTM freeware , developed as an algorithm for calculation radiance of the scattering and absorbing Earth's atmosphere and transmissivity computations within the waveband range from UV to radar.  
		\item [*] The \textit{Community Radiative Transfer Model}, \href{http://www.jcsda.noaa.gov/projects_crtm.php}{CRTM} is designed by the  (\href{http://www.jcsda.noaa.gov/}{JCSDA, NASA}) and includes Surface Emissivity/Reflectivity Models, Cloud Absorption/Scattering Model and Gaseous Absorption Model.
		\item [*] An \textit{Automatized Atmospheric Absorption Atlas Operational Release} \href{http://www.noveltis.fr/4AOP/}{(4AOP)} is a French RTM freeware, designed in cooperation by \href{http://www.noveltis.fr/fr/index.php}{Noveltis} and \href{http://www.cnes.fr/}{CNES} (the French Space Agency). Being a fine program which enables accurate computations of the radiance, defining other emissivity functions and other set up, it is however, particularly designed for the infrared waves (thus, not quite suitable in our case). To the advantages of this freeware we would count its user friendly interface and open source accessibility for the academic purposes. 
	\end{itemize}
For our purpose we have chosen the \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{\textit{WASI Water Color Simulator}} RTM software, due to its effectiveness, adaptability for the Mediterranean environment, open source availability, coverage of necessary wavebands and clear, user friendly interface enabling us to adjust various environmental parameters.

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03} \label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, markers, GPS and other equipment. The seagrass sampling is
taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}. \\
Other methods of seagrass \textit{in-situ} monitoring are based on
using the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then
collect the return echoes for the analysis, yet this requires specialized equipment and is mostly used in
the depth waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that they are not effective at identifying the composition or even the presence of
submerged aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}, because
initially acoustic methods were intended for the bathymetric surveying, i.e. they are not adjusted for
the benthic habitat discrimination. The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated GIS approach, as, for example, reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.

\subsection[Application of the remote sensing data]{Application of the remote sensing data towards seagrass mapping}
Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, GPS data, remote sensing and SCUBA-based fieldwork measurements.
SCUBA-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography with a GPS is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}. However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. Integrated together with GIS vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:13}
\end{wrapfigure}

However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Mellors09a}\label{Mellors09a}.
Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance.

@@

Spectral measurements of the target objects are made
by means of the radiometers, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.
Studies of spectral reflectance of the different seagrass species comparing to the spectra of sand and
other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (LAI) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.

\begin{table}[htbp]
	\caption{Characteristics of selected ocean-colour sensors}
	\begin{center}
		\rowcolors{1}{OldLace}{FloralWhite}
		\begin{tabular}{|p{2cm}|p{20mm}|p{15mm}|p{17mm}|p{15mm}|p{15mm}|p{15mm}|}
			\hline\hline
			\textbf{Satellite Sensor:} & {\textbf{Landsat}} & {\textbf{SPOT}} & {\textbf{ENVISAT}} & {\textbf{Terra}} & {\textbf{GeoEye}} & {\textbf{Nimbus 7}} \\ \hline\hline
				\textbf{Origin} & USA & France & Europe (ESA) & USA (NASA) & USA & USA \\ \hline
				\textbf{Instrument} & \textit{TM, ETM+} & \textit{Pan, XS} & \textit{MERIS} & \textit{MODIS} & \textit{SeaWiFS} & \textit{CZCS}\\ \hline
			\textbf{Number of channels} & 8 & 6 & 15 & 36 & 8 & 5 \\ \hline
			\textbf{Wavelength coverage} & 185-1700 & 500-1750 & 412.5- 900 & 469-1640 & 412-865 & 443-750 \\ \hline
			\textbf{Ground resolution (nadir), m} & 15-pan,30(MS),60 (TIR) & 2.5, 5, 10,20 & 1.2 km/300 m & 1.0 km & 1.13 km & 825 m \\ \hline
			\textbf{Band Width (nm)} & 0.45-10.4 & 10,20 & 2.5,7.5,9,10 14,20 & 0.18,0.3,0.5,10 - 50 & 20,40 & 20,100 \\ \hline
			\textbf{Launched} & 1972, 1999 & 1986 & 2002 & 1999 & 1997 & 1978 \\ \hline
			\textbf{Recurrent period, days} & 16 & 24 & 35 & 16 & 16 & 16 \\ \hline
		\end{tabular}
	\end{center}
	\label{tab:1}
\end{table}

Spectral reflectance of the seagrass  (Fig.\ref{fig:13}) is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason for marine mapping of submerged vegetation
and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).
In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.

@@

The Landsat TM and Landsat ETM+ data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
Landsat TM towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.
Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. 
The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. Thus,
the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, Landsat TM, and CASI, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and CASI, while Landsat TM has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that CASI enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.

@@

\subsection{Fieldwork data collection}
Seagrass sampling has been performed at two research stations at Crete Island - Ligaria beach
($36\,^{\circ}20\textquoteright N, 22\,^{\circ}59\textquoteright E$) and Xerocampos ($35\,^{\circ}12\textquoteright N, 26\,^{\circ}18\textquoteright E$), at depths lesser than 4 meters.
